Applying Renormalization Group Theory to the Square Well Liquid

### 2014-05-31 papers/thesis-roth/project.tex
fix typo in project.tex Tested on bennet. Took 0:18.45.
project report Tested on elliot. Took 0:19.23.

\documentclass[letterpaper,twocolumn,amsmath,amssymb,prb]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

% special values for n
\newcommand{\npart}{\ensuremath{n_{particular}}}
\newcommand{\nliq}{\ensuremath{n_{liquid}}}
\newcommand{\nvap}{\ensuremath{n_{vapor}}}

@@

@@

Near the critical point of a liquid, there are density fluctuations at
all length scales. In this regime, mean field theories break down. As
the critical point is approached, there becomes less and less of a
difference in free energy of the liquid and vapor states; this causes
difficulty in modeling liquid-vapor coexistence. This work develops a
method for calculating liquid-vapor coexistence, and applies
renormalization group theory to the square well liquid.

@@

Gasses have been accurately modeled since the 19$^{th}$
century\cite{Lederman92} and the kinetic theory of gasses. With the
advent of quantum mechanics in the early 20$^{th}$ century, we have
had a pretty good grasp on solids. A robust model of liquids continues
to be elusive. Liquids are not periodic structures like solids, but
neither are they disperse, as are gasses.

@@

One particularly troublesome regime is around the so-called ``critical
point'' of a liquid. At the appropriate density and temperature, the
distinction between liquid and vapor dissapears. Beyond this point,
you have a ``supercritical fluid.'' Supercritical fluids are used in
many industrial processes\cite{Perrut00}. For example (and of personal
interest), supercritical fluids are used in coffee decaffeination and
the extraction of hops resins for brewing beer.

@@

We use statistical mechanics to study liquids; this requires large
sample sizes, long simulation times, or some combination of the
two. Monte Carlo simulations are exact, if you allow them to run
``long enough.'' Our models are ultimately compared to these
simulations, but Monte Carlo simulations are otherwise outside the
scope of this work. The Roundy group uses classical Density-funtional
Theory (DFT) to calculate free energies and probability distributions
of bulk liquids. The end goal of this work is to investigate into
wheather or not implementing renormalization group theory is
worthwhile to the Roundy research group.

In section~\ref{sec:theory}, I present some of the main theories I
deal with: renormalization group theory, thermodynamic perturbation
theory and the square well liquid free energy. In
section~\ref{sec:methods} I discuss the the methods used. I cover the
calculation of liquid-vapor coexistence curves in some detail, as
caluclating coexistence near the critical point is tricky.

@@

We start by dividing the total volume into cells of volume $V_i$. We
assume that within a cell $n(\rr)$ varies only slightly, and that the
cells do not interact with one another. This is refered to as Wilson's
phase cell method.\cite{Ramana12}

Then consider a baseline free enrgy $F_0(T,n)$. We assume this
baseline accounts for all fluctuations of wavelength $\lambda <
\lambda_0$. We then double the size of our phase cell and add a
correction to the free energy $\delta F_1(T,n)$. This correction
accounts for fluctuations with a wavelength $\lambda =
2\lambda_0$. The new free energy is
\begin{align}
  F_1(T,n) &= F_0(T,n) + \delta F_1(T,n)
\end{align}
Then double the length scale again; $\delta F_2(T,n)$ accounts for
fluctionations with $\lambda = 4\lambda_0$ and
\begin{align}
  F_2(T,n) &= F_1(T,n) + \delta F_2(T,n)
\end{align}
As we continue this process, the $i^{th}$ correction accounts for
fluctuations with $\lambda = 2^i\lambda_0$, and
\begin{align}
  F_i(T,n) &= F_{i-1}(T,n) + \delta F_i(T,n)
\end{align}
In the end, your free energy is
\begin{align}
  F(T,n) = F_0(T,n) + \sum_{i=1}^\infty\delta F_i(T,n) + \Fattr
\end{align}

Where $\Fattr$ is the contribution due to the longest-wavelength
density fluctuations. At this stage, I do not entirely understand why
Forte splits up $\Fattr$.

The $\detla F_i(T,n)$ term is a difference in free energies:
\begin{align}
  \delta F_i(T,n) &= -\kT\ln Z - \left( -\kT\ln Z^* \right) \\
  &= -\kT\ln\left[ \frac{Z_D(T,n)}{Z_D^*(T,n} \right]
\end{align}
Here, $Z$ is the partition function for the entire system, and $Z^*$ is the partition function of a reference system consisting only of the term $\FbarD$. Forte gives these terms as
\begin{align}
  Z_D(T,n) &= \int dx\, \exp\left[ -\frac{V_D}{\kT}\left( \FbarD + \bar{U}_D(T,n \right) \right] \\
  Z_D^*(T,n) &= \int dx\, \exp\left[ -\frac{V_D}{\kT}\FbarD \right]
\end{align}
where $x$ is the amplitude of density fluctuations. $\FbarD$ and $\UbarD$ are formulated similar to averages:
\begin{align}
  \FbarD &= \frac{F_{i-1}(T,n+x) + F_{i-1}(T,n-x)}{2} - F_{i-1}(T,n) \\
  \UbarD &= \frac{U(T,n+x) + U(T,n-x)}{2} - U(T,n)
\end{align}
Finally, $U(T,n)$ is given by Forte as
\begin{align}
  U(T,n) = &{} g^\text{HS}(n_{eff};\sigma)\alpha - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha\omega^2 \nonumber \\
  &{} + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha\gamma \label{eqn:URG} % URG! It sounds like it hurt! :P
\end{align}
with
\begin{align}
  \alpha &= \frac{2}{3}\pi\epsilon\sigma^3(\lambda^3 - 1) \\
  \omega^2 &= \frac{1}{5}\sigma^2\frac{\lambda^5 - 1}{\lambda^3 - 1} \\
  \gamma &= \frac{1}{70}\sigma^4\frac{\lambda^7 - 1}{\lambda^3 - 1}
\end{align}
The pair correlation function $g(n,r)$ tells us how likely we are to find another hard sphere some distance away, given some density. In equation~\ref{eqn:URG} we are considering this quantity at contact ($r = \sigma$) with some effective density.

@@

Thermodynamic perturbation theory (TPT) gives a good description of
simple liquids, as long as you are far from the critical point---which
is most of where we calculate. TPT is a rigorous power series
expansion in a small parameter; as long as the parameter is small, the
theory is valid. This discussion follows primarily from \textit{Theory
  of Simple Liquids}, by Hansen and McDonald\cite{Hansen06}.

@@

We consider this as our referecne, and take a square well attraction as a perturbation:
\begin{align}
  v_\lambda(\1,\2) &= v_\text{HS}(\1,\2) + \lambda w^\text{SW}(\1,\2) \label{eqn:small-perturbation}
\end{align}
where
\begin{align}
  w^\text{SW}(\1,\2) &= - \varepsilon \text{ for } \sigma < r < \lambda\sigma
\end{align}

The total potential energy of our system and its derivatives with resepect to $\lambda$---which become important later---are
\begin{align}
  \widetilde{V} &= \sum_{i=1}^N\sum_{j>i}^N v_\lambda(\mathbf{r}_i,\mathbf{r}_j) \\ % \label{eq:VN} \\
  \hookrightarrow \frac{\partial \widetilde{V}}{\partial\lambda} &= \sum_{i=1}^N\sum_{j>i}^N w^\text{SW}(\mathbf{r}_i,\mathbf{r}_j) \nonumber \\
  &\equiv W_N \\
  \hookrightarrow \frac{\partial^2 \widetilde{V}}{\partial\lambda^2} &= 0
\end{align}

The \emph{configuration integral} (the potential energy component of the partition function) and its derivative with respect to $\lambda$ are
\begin{align}
  Z_\lambda &= \int d\rr^N\, e^{-\beta \widetilde{V}(\lambda)} \\
  \hookrightarrow \frac{\partial Z_\lambda}{\partial\lambda} &=  \int d\rr^N\, \frac{\partial}{\partial\lambda}e^{-\beta \widetilde{V}(\lambda)} \\
  &= \int d\rr^N\, -\beta\frac{\partial \widetilde{V}}{\partial\lambda}e^{-\beta \widetilde{V}(\lambda)} \\
  &= \int d\rr^N\, -\beta W_N e^{-\beta \widetilde{V}(\lambda)}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free energy) and its derivative are
\begin{align}
  F &= -\kT\ln\left( \frac{Z_\lambda}{V^N} \right) \\
  \hookrightarrow \frac{\partial F}{\partial\lambda} &= \frac{V^N}{Z_\lambda}\frac{\partial}{\partial\lambda}\left( \frac{Z_\lambda}{V^N} \right) \\
  &= -\kT\left( \frac{1}{Z_\lambda}\frac{\partial Z_\lambda}{\partial\lambda} \right) \\
  &= \frac{1}{Z_\lambda} \int d\rr^N\, W_N e^{-\beta \widetilde{V}(\lambda)}\\
  &= \left\langle W_N \right\rangle_\lambda \label{eqn:dfdlambda}
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and
$\left\langle W_N \right\rangle_\lambda$ is the ensemble
average of $W_N$, with the subscript indicating some given
value of $\lambda$.

\fixme{The following could probably go someplace else (is it even necessary?):
  We can simplify this even further by noting
  \begin{align}
    \left\langle W_N \right\rangle &= \left\langle \sum_i\sum_jw(\mathbf{r}_i,\mathbf{r}_j) \right\rangle \\
    &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle
    \intertext{which gives us}
    \frac{\partial F}{\partial\lambda} &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle_\lambda
\end{align}}

Note that equation~\ref{eqn:dfdlambda} is exact; now take a series expansion:
\begin{align}
  \left\langle W_N \right\rangle_\lambda &= \left\langle W_N\right\rangle_{\lambda = 0} + (\lambda)\frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda}\bigg|_{\lambda = 0} + \mathcal{O}(\lambda^2) \label{eqn:Wn-expansion}
\end{align}
The derivative term is
\begin{widetext}
  \begin{align}
    \frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda} &= \frac{\partial}{\partial\lambda}\left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}}\right) \\
    &= \frac{1}{Z_\lambda}\frac{\partial}{\partial\lambda}\left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\frac{\partial}{\partial\lambda}Z_\lambda \\
    &= \frac{1}{Z_\lambda}\int d\rr^N\, \left( e^{-\beta \widetilde{V}}\frac{\partial}{\partial\lambda}W_N + W_N\frac{\partial}{\partial\lambda}e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\int d\rr^N\, \frac{\partial}{\partial\lambda}e^{-\beta \widetilde{V}} \\
  \end{align}
\end{widetext}

Noting that $\partial W_N/\partial\lambda = 0$, we have
\begin{widetext}
  \begin{align}
    \frac{\partial}{\partial\lambda}\left\langle W_N \right\rangle_{\lambda} &= \frac{1}{Z_\lambda}\int d\rr^N (-\beta) W_N^2 e^{-\beta \widetilde{V}} - \frac{1}{Z_\lambda^2}\left(\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\left( \int d\rr^N\, (-\beta)W_Ne^{-\beta \widetilde{V}} \right) \\
    &= -\beta\left[ \frac{1}{Z_\lambda}\int d\rr^N\, W_N^2e^{-\beta \widetilde{V}} - \left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)^2 \right] \\
    &= -\beta\left[ \left\langle W_N^2 \right\rangle_\lambda - \left\langle W_N \right\rangle_\lambda^2 \right] \\
    &\equiv -\beta\sigma_\lambda^2
  \end{align}
\end{widetext}

Combining this result with equation~\ref{eqn:Wn-expansion} gives us
\begin{align}
  \left\langle W_N \right\rangle_\lambda &= \left\langle W_N\right\rangle_{\lambda = 0} - \beta\lambda\sigma_{\lambda=0}^2 + \mathcal{O}(\lambda^2) \label{eqn:Wn-expansion-simplified}
\end{align}
and we can insert this into equation~\ref{eqn:dfdlambda}:
\begin{align}
  \frac{\partial F}{\partial\lambda} &= \left\langle W_N\right\rangle_{\lambda = 0} - \beta\lambda\sigma_{\lambda=0}^2 + \mathcal{O}(\lambda^2)
\end{align}

We can integrate this expression over $F\in[F_0,F]$ and $\lambda\in[0,1]$ to obtain
\begin{align}
  F - F_0 &= (1-0)\left\langle W_N \right\rangle_\lambda - \frac{1}{2}\beta(1 - 0)^2\sigma_{\lambda=0}^2 + \mathcal{O}(\beta^2) \nonumber \\
  \hookrightarrow F &= F_0 + \left\langle W_N \right\rangle_\lambda - \frac{1}{2}\beta\sigma_{\lambda=0}^2 + \mathcal{O}(\beta^2) \label{eqn:F-inTermsW}
\end{align}
Multiply by $\beta$ and we have the \emph{high-temperature expansion}, first derived by
Zwanzig:\cite{Zwanzig54}
\begin{align}
  \beta F &= \beta F_0 + \beta \left\langle W_N \right\rangle_{\lambda=0} - \beta^2\lambda\sigma_{\lambda=0}^2 + \mathcal{O}(\beta^3)
\end{align}

Zwanzig further showed that the $n^{th}$ term could be written in terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle W_N\right\rangle_{\lambda=0} \right]^\nu \right\rangle_{\lambda=0}$ with $\nu \leq n$. This requires us to know the form of higher-order correlation functions of the reference system, which can be quite complicated. In the next section we will see a simple aproximation for the $n=2$ term.

@@

\subsection{Square well liquid free enrgy}\label{subsec:SW}
For the baseline, this work is based on the square well liquid free energy:\cite{Hughes13}
\begin{align}
  f_0(T,n) &= \fid + \fhs + \left( \fdisp - n a_1(n) \right) \ .
\end{align}
The long-range attractive term, $\fattr$ is given by
\begin{align}
  \fattr &= n a_1(n) \ .
\end{align}
$\fid$ is the ideal gas free energy, $\fhs$ is the hard sphere
repulsion, and $\fdisp$ is the attraction. The function $a_1(n)$ is
the first term from a high-temperature perturbation expansion,
described in further detail in sections~\ref{subsec:TPT}
and~\ref{sub2sec:disp}. Each function is outlined below.

@@

\subsubsection{Hard sphere repulsion}\label{sub2sec:HS}
Hard sphere repulsive forces are dealt with by the White Bear version
of Fundamental-Mearuse Theory.\cite{Roth02} The hard sphere excess
free energy is given in the homogeneous case as
\begin{align}
  \fhs &= \kT\left( \Phi_1(n) + \Phi_2(n) + \Phi_3(n) \right)
\end{align}
The $\Phi_j(n)$ terms are given as
\begin{align}
  \Phi_1(n) &= -n_0(n)\ln(1 - n_3(n)) \\
  \Phi_2(n) &= \frac{n_1(n)n_2(n)}{1 - n_3(n)} \\
  \Phi_3(n) &= n_2^3(n)\left( \frac{n_3(n) + (1 - n_3(n))^2\ln(1 - n_3(n))}{36\pi n_3^2(n)(1 - n_3(n))^2} \right)
\end{align}
The \emph{fundamental measure densities} are
\begin{align}
  n_0(n) &= \frac{n_2}{4\pi R^2} \nonumber \\
  &= n \\
  n_1(n) &= \frac{n_2}{4\pi R} \nonumber \\
  &= nR \\
  n_2(n) &= n4\pi R^2 \\
  n_3(n) &= n\frac{4}{3}\pi R^3 \nonumber \\
  &= \eta
\end{align}
where $\eta$ is the packing fraction.

@@

For this work, I use $\lambda_\text{SW} = 1.5$.

>>>>

\documentclass[letterpaper,twocolumn,amsmath,amssymb,prb]{revtex4-1}
\usepackage[demo]{graphicx}% Include figure files. When building at home, use the option [demo] to supress actual need of figures
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

% special values for n
\newcommand{\npart}{\ensuremath{n_{part}}}
\newcommand{\nliq}{\ensuremath{n_{liquid}}}
\newcommand{\nvap}{\ensuremath{n_{vapor}}}

@@

% lambda
\newcommand{\lambdaSW}{\ensuremath{\lambda_\text{SW}}}
\newcommand{\lambdaRG}{\ensuremath{\lambda_\text{RG}}}

@@

Near the critical point of a liquid, there are density fluctuations at
all length scales. In this regime, mean field theories break down. \fixme{re-word the next sentence} As
the critical point is approached, there becomes less and less of a
difference in free energy of the liquid and vapor states; this causes
difficulty in modeling liquid-vapor coexistence. This work develops a
method for calculating liquid-vapor coexistence, and applies
renormalization group theory to the square well liquid.

@@

Gasses have been accurately modeled since the 19$^{th}$
century\cite{Lederman92} and formulation of the kinetic theory of gasses. With the
advent of quantum mechanics in the early 20$^{th}$ century, we have
had a pretty good grasp on solids. A robust model of liquids continues
to be elusive. Liquids are not periodic structures like solids, but
neither are they disperse, as are gasses.

@@

One particularly troublesome regime is around the critical
point of a liquid. At the appropriate density and temperature, the
distinction between liquid and vapor dissapears. Beyond this point,
you have a ``supercritical fluid.'' Supercritical fluids are used in
many industrial processes.\cite{Perrut00} For example (and of personal
interest), supercritical fluids are used in coffee decaffeination and
the extraction of hops resins for brewing beer.

It is the aim of renormalization group theory to tackle the critical
point. The reason that calculations are dificult near the critical
point is that there are large-scale fluctuations of the density. In
fact, there are density fluctuations at \emph{all} length
scales, which is why mean field theories fail. Renormalization group theory is an iterative procedure that
was developed in th 1950's, and Kenneth Wilson's work with the theory
in the 1970's won him the 1982 Nobel Prize in physics.\cite{Nobel82}

\fixme{Transition here?}

@@

We use statistical mechanics to study liquids; this requires large
sample sizes, long simulation times, or some combination of the
two. Monte Carlo simulations are exact, if you allow them to run
``long enough.'' The Roundy group uses classical Density-funtional
Theory (DFT) to calculate free energies and probability distributions
of bulk liquids. These calculations are ultimately compared to Monte Carlo simulations.

The end goal of this work is to investigate into whether or not
implementing renormalization group theory is worthwhile to the Roundy
Research Group. Many other groups use renormalization group theory,
but none of them tell the iteration depth used in their
calculations.\cite{Forte11,Ramana12,Ghobadi13,Wang13,Zhao13} If only
few iterations are required, then this would certainly be an effective
tool to use. However, if many iterations are required, the
computational cost could be unwieldy.

In section~\ref{sec:theory}, I present some of the main theories I
deal with: renormalization group theory, thermodynamic perturbation
theory and the square well liquid free energy. In
section~\ref{sec:methods} I discuss the the methods used. I cover the
calculation of liquid-vapor coexistence curves in some detail, as
caluclating coexistence near the critical point is tricky. The
results---such as they are---are covered in section~\ref{sec:results}.

@@

We start by dividing the total volume into cells of volume $V_i$. We
assume that within a cell the number density $n(\rr)$ varies only slightly, and that the
cells do not interact with one another. This comes from Wilson's
phase cell method.\cite{Ramana12}

Then consider a baseline free enrgy $F_0(T,n)$, usually based on a
mean field theory. We assume this baseline accounts for all
fluctuations of wavelength $\lambdaRG < \lambda_0$. We then
double the size of our phase cell and add a correction to the free
energy $\delta F_1(T,n)$. This correction accounts for fluctuations
with a wavelength $\lambdaRG = 2\lambda_0$. The new free
energy, $F_1(T,n)$, is
\begin{align}
  F_1(T,n) &= F_0(T,n) + \delta F_1(T,n)
\end{align}
Then double the length scale again; $\delta F_2(T,n)$ accounts for
fluctionations with $\lambdaRG = 4\lambda_0$ and
\begin{align}
  F_2(T,n) &= F_1(T,n) + \delta F_2(T,n)
\end{align}
As we continue this process, the $i^{th}$ correction accounts for
fluctuations with $\lambdaRG = 2^i\lambda_0$, and
\begin{align}
  F_i(T,n) &= F_{i-1}(T,n) + \delta F_i(T,n)
\end{align}
In the end, your free energy is
\begin{align}
  F(T,n) = F_0(T,n) + \sum_{i=1}^\infty\delta F_i(T,n) + \Fattr
\end{align}

Where $\Fattr$ is the contribution due to the longest-wavelength
density fluctuations. At this stage, I do not entirely understand why
Forte separates out $\Fattr$.

The $\delta F_i(T,n)$ term is a difference in free energies:
\begin{align}
  \delta F_i(T,n) &= -\kT\ln Z - \left( -\kT\ln Z^* \right) \nonumber \\
  &= -\kT\ln\left[ \frac{Z(T,n)}{Z^*(T,n)} \right]
\end{align}
Here, $Z$ is the partition function for the entire system, and $Z^*$
is the partition function of a reference system consisting only of the
term $\FbarD$. Forte gives these terms as
\begin{align}
  Z(T,n) &= \int dx\, \exp\left[ -\frac{V}{\kT}\left( \FbarD + \bar{U}(T,n \right) \right] \label{eqn:Z} \\
  Z^*(T,n) &= \int dx\, \exp\left[ -\frac{V}{\kT}\FbarD \right] \label{eqn:Zstar}
\end{align}
where $x$ is the amplitude of density fluctuations. $\FbarD$ and $\UbarD$ are formulated similar to averages:
\begin{align}
  \FbarD &= \frac{F_{i-1}(T,n+x) + F_{i-1}(T,n-x)}{2} - F_{i-1}(T,n) \\
  \UbarD &= \frac{U(T,n+x,\lambdaRG) + U(T,n-x,\lambdaRG)}{2} - U(T,n,\lambdaRG)
\end{align}
Finally, $U(T,n,\lambdaRG)$ is given by Forte as
\begin{align}
  U(T,n,\lambdaRG) = &{} g^\text{HS}(n_{eff};\sigma)\alpha(\lambdaRG) - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha(\lambdaRG)\omega^2(\lambdaRG) \nonumber \\
  &{} + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha(\lambdaRG)\gamma(\lambdaRG) \label{eqn:URG} % URG! It sounds like it hurt! :P
\end{align}
with
\begin{align}
  \alpha(\lambdaRG) &= \frac{2}{3}\pi\epsilon\sigma^3(\lambdaRG^3 - 1) \\
  \omega^2(\lambdaRG) &= \frac{1}{5}\sigma^2\frac{\lambdaRG^5 - 1}{\lambdaRG^3 - 1} \\
  \gamma(\lambdaRG) &= \frac{1}{70}\sigma^4\frac{\lambdaRG^7 - 1}{\lambdaRG^3 - 1}
\end{align}
The pair correlation function $g(n,r)$ tells us how likely we are to
find another hard sphere some distance away, given some density. In
equation~\ref{eqn:URG} we are considering this quantity at contact ($r
= \sigma$) with an effective density $n_{eff}$.

\fixme{Transition?}

@@

Thermodynamic perturbation theory (TPT) gives a good description of
simple liquids, as long as you are far from the critical point---which
is most of the time. TPT is a rigorous power series
expansion in a small parameter; as long as the parameter is small, the
theory is valid. This discussion follows primarily from \textit{Theory
  of Simple Liquids}, by Hansen and McDonald\cite{Hansen06}.

@@

We consider this as our referecne, and take a square well attraction as a perturbation:
\begin{align}
  w_\text{SW}(\1,\2) &= - \varepsilon \text{ for } \sigma < r < \lambdaSW\sigma
\end{align}
so that we have
\begin{align}
  v_\lambda(\1,\2) &= v_\text{HS}(\1,\2) + \lambdaSW w_\text{SW}(\1,\2) \label{eqn:small-perturbation}
\end{align}

The total potential energy of our system and its derivatives with
resepect to $\lambdaSW$ (which become important later) are
\begin{align}
  \widetilde{V} &= \sum_{i=1}^N\sum_{j>i}^N v_\lambda(\mathbf{r}_i,\mathbf{r}_j) \\ % \label{eq:VN} \\
  \frac{\partial \widetilde{V}}{\partial\lambdaSW} &= \sum_{i=1}^N\sum_{j>i}^N w_\text{SW}(\mathbf{r}_i,\mathbf{r}_j) \nonumber \\
  &\equiv W_N \\
  \frac{\partial^2 \widetilde{V}}{\partial\lambdaSW^2} &= 0
\end{align}

The \emph{configuration integral} (the potential energy component of
the partition function) and its derivative with respect to $\lambdaSW$
are
\begin{align}
  Z_\lambda &= \int d\rr^N\, e^{-\beta \widetilde{V}} \\
  \frac{\partial Z_\lambda}{\partial\lambdaSW} &=  \int d\rr^N\, \frac{\partial}{\partial\lambdaSW}\left( e^{-\beta \widetilde{V}} \right) \nonumber \\
  &= \int d\rr^N\, -\beta\frac{\partial \widetilde{V}}{\partial\lambda}e^{-\beta \widetilde{V}} \nonumber \\
  &= \int d\rr^N\, -\beta W_N e^{-\beta \widetilde{V}}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free
energy) and its derivative are
\begin{align}
  F &= -\kT\ln\left( \frac{Z_\lambda}{V^N} \right) \\
  \frac{\partial F}{\partial\lambdaSW} &= \frac{V^N}{Z_\lambda}\frac{\partial}{\partial\lambdaSW}\left( \frac{Z_\lambda}{V^N} \right) \nonumber \\
  &= -\kT\left( \frac{1}{Z_\lambda}\frac{\partial Z_\lambda}{\partial\lambdaSW} \right) \nonumber \\
  &= \frac{1}{Z_\lambda} \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \label{eqn:dfdlambda_Z}\\
  &= \left\langle W_N \right\rangle_{\lambdaSW} \label{eqn:dfdlambda}
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$ before equation~\ref{eqn:dfdlambda_Z}). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and $\left\langle W_N
\right\rangle_{\lambdaSW}$ is the ensemble average of $W_N$, with the
subscript indicating some value of $\lambdaSW$.

%% \fixme{The following could probably go someplace else (is it even necessary?):
%%   We can simplify this even further by noting
%%   \begin{align}
%%     \left\langle W_N \right\rangle &= \left\langle \sum_i\sum_jw(\mathbf{r}_i,\mathbf{r}_j) \right\rangle \\
%%     &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle
%%     \intertext{which gives us}
%%     \frac{\partial F}{\partial\lambda} &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle_\lambda
%% \end{align}}

Note that equation~\ref{eqn:dfdlambda} is exact; now take a series expansion:
\begin{widetext}
\begin{align}
  \left\langle W_N \right\rangle_{\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} + (\lambdaSW)\frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW}\bigg|_{\lambdaSW = 0} + \mathcal{O}(\lambdaSW^2) \label{eqn:Wn-expansion}
\end{align}
%% \end{widetext}
The derivative term is
%% \begin{widetext}
  \begin{align}
    \frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW} &= \frac{\partial}{\partial\lambdaSW}\left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}}\right) \nonumber \\
    &= \frac{1}{Z_\lambda}\frac{\partial}{\partial\lambdaSW}\left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\frac{\partial}{\partial\lambdaSW}Z_\lambda \nonumber \\
    &= \frac{1}{Z_\lambda}\int d\rr^N\, \left( e^{-\beta \widetilde{V}}\frac{\partial}{\partial\lambdaSW}W_N + W_N\frac{\partial}{\partial\lambdaSW}e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\int d\rr^N\, \frac{\partial}{\partial\lambdaSW}e^{-\beta \widetilde{V}}
  \end{align}
%% \end{widetext}

Noting that $\partial W_N/\partial\lambda = 0$, we have
%% \begin{widetext}
  \begin{align}
    \frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW} &= \frac{1}{Z_\lambda}\int d\rr^N (-\beta) W_N^2 e^{-\beta \widetilde{V}} - \frac{1}{Z_\lambda^2}\left(\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\left( \int d\rr^N\, (-\beta)W_Ne^{-\beta \widetilde{V}} \right) \nonumber \\
    &= -\beta\left[ \frac{1}{Z_\lambda}\int d\rr^N\, W_N^2e^{-\beta \widetilde{V}} - \left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)^2 \right] \nonumber \\
    &= -\beta\left[ \left\langle W_N^2 \right\rangle_{\lambdaSW} - \left\langle W_N \right\rangle_{\lambdaSW}^2 \right] \nonumber \\
    &\equiv -\beta\sigma_{\lambdaSW}^2
  \end{align}
\end{widetext}

Combining this result with equation~\ref{eqn:Wn-expansion} gives us
\begin{align}
  \left\langle W_N \right\rangle_{\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2) \label{eqn:Wn-expansion-simplified}
\end{align}
and we can insert this into equation~\ref{eqn:dfdlambda}:
\begin{align}
  \frac{\partial F}{\partial\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2)
\end{align}

Now integrate this expression over $F\in[F_0,F]$ (\textsc{N.B.}~this is different than the RGT baseline free energy $F_0(T,n)$ in section~\ref{subsec:RGT}) and $\lambda\in[0,1]$ to obtain
\begin{align}
  \int_{F_0}^F\, dF &= \int_0^1 d\lambdaSW\, \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2) \nonumber \\
  \intertext{\fixme{Do I need that first line?}}
  \hookrightarrow F - F_0 &= (1-0)\left\langle W_N \right\rangle_{\lambdaSW} - \frac{1}{2}\beta(1 - 0)^2\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^2) \nonumber \\
  \hookrightarrow F &= F_0 + \left\langle W_N \right\rangle_{\lambdaSW} - \frac{1}{2}\beta\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^2) \label{eqn:F-inTermsW}
\end{align}
Multiply by $\beta$ and we have the \emph{high-temperature expansion}, first derived by
Zwanzig:\cite{Zwanzig54}
\begin{align}
  \beta F &= \beta F_0 + \beta \left\langle W_N \right\rangle_{\lambdaSW=0} - \beta^2\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^3)
\end{align}

Zwanzig further showed that the $n^{th}$ term could be written in
terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle
  W_N\right\rangle_{\lambdaSW=0} \right]^\nu \right\rangle_{\lambdaSW=0}$
with $\nu \leq n$. This requires us to know the form of higher-order
correlation functions of the reference system, which can be quite
complicated. In the next section we will see a simple aproximation for
the $n=2$ term.

@@

\subsection{Square well liquid free enrgy}\label{subsec:SW}
\fixme{I need to either note here that I use F/V (see start of section~\ref{sec:methods}), or multiply every term in this section by V}
For the baseline, this work uses the square well liquid free energy:\cite{Hughes13}
\begin{align}
  f_0(T,n) &= \fid + \fhs + \left( \fdisp - n a_1(n) \right) \ .
\end{align}
$\fid$ is the ideal gas free energy, $\fhs$ is the hard sphere
repulsion, and $\fdisp$ is the attraction. The function $a_1(n)$ is
the first term from a high-temperature perturbation expansion,
described in further detail in sections~\ref{subsec:TPT}
and~\ref{sub2sec:disp}. Each function is outlined below.

@@

\subsubsection{Hard sphere repulsion}\label{sub2sec:HS}
Hard sphere repulsive forces are dealt with by the White Bear version
of Fundamental-Mearuse Theory.\cite{Roth02} The hard sphere excess
free energy is given in the homogeneous case as
\begin{align}
  \fhs &= \kT\left( \Phi_1(n) + \Phi_2(n) + \Phi_3(n) \right)
\end{align}
The $\Phi_j(n)$ terms are given as
\begin{align}
  \Phi_1(n) &= -n_0(n)\ln(1 - n_3(n)) \\
  \Phi_2(n) &= \frac{n_1(n)n_2(n)}{1 - n_3(n)} \\
  \Phi_3(n) &= n_2^3(n)\left( \frac{n_3(n) + (1 - n_3(n))^2\ln(1 - n_3(n))}{36\pi n_3^2(n)(1 - n_3(n))^2} \right)
\end{align}
The \emph{fundamental measure densities} are
\begin{align}
  n_0(n) &= \frac{n_2}{4\pi R^2} \nonumber \\
  &= n \\
  n_1(n) &= \frac{n_2}{4\pi R} \nonumber \\
  &= nR \\
  n_2(n) &= n4\pi R^2 \\
  n_3(n) &= n\frac{4}{3}\pi R^3 \nonumber \\
  &= \eta \label{eqn:packingFrac}
\end{align}
where $\eta$ is the packing fraction.

@@

For this work, I use $\lambdaSW = 1.5$.
